---
title: Uncertainty Estimation for High-dimensional Nonparametric Forecasting
author: Nuwani Palihawadana
date: 01 July 2025
titlegraphic: bg-13.png
titlecolor: white
toc: false
format:
  presentation-beamer:
    pdf-engine: pdflatex
    keep-tex: true
    fig-width: 8
    fig-height: 4.3
    template-partials:
      - before-title.tex
highlight-style: tango
execute:
  echo: false
  message: false
  warning: false
  cache: false
---

```{r}
#| label: load-packages
#| echo: false
library(knitr)
library(ggplot2)
library(dplyr)
library(tidyr)
library(tsibble)
library(lubridate)
library(cowplot)
library(kableExtra)
library(feasts)
library(tikzDevice)
library(shape)
```

## Nonparametric forecasting

\begin{block}{}
$$
 y_t = f(\bm{a}_{t}, \bm{a}_{t-1}, \dots,\bm{a}_{t-s}, y_{t-1},\dots,y_{t-k}) + \varepsilon_{t},
$$
\end{block}

\begin{itemize}
  \item \small \color{black} $y_{t}$ -- \color{violet} observed response at time $t$
  \item \small \color{black} $f$ -- \color{violet} arbitrary function
  \item \small \color{black} $\bm{a}_{t}$ -- \color{violet} vector of exogenous variables at time $t$
  \item \small \color{black} $\varepsilon_{t}$ -- \color{violet} stationary error with mean zero and constant variance $\sigma^2$. 
\end{itemize}

\pause

We also define:

\begin{itemize}
  \item \small \color{black} $\bm{z}_{t} = (y_{t}, \bm{x}_{t})$ -- \color{violet} observation at time $t$
  \item \small \color{black} $\bm{x}_{t} = (\bm{a}_{t}, \bm{a}_{t-1}, \dots,\bm{a}_{t-s}, y_{t-1},\dots,y_{t-k})$ -- \color{violet} vector of all predictors at time $t$.
  \item \small \color{black} $\hat{y}_{t} = \hat{f}(\bm{x}_{t})$  -- \color{violet} estimated model
  \item \small \color{black} $e_{t} = y_{t} - \hat{y}_{t}$  -- \color{violet} model residuals
\end{itemize}



## Sparse multiple index (SMI) model

\begin{block}{}
$$y_{i} = \beta_{0} + \sum_{j = 1}^{p}g_{j}(\bm{\alpha}_{j}^{T}\bm{x}_{ij}) + \sum_{k = 1}^{d}f_{k}(w_{ik}) + \bm{\theta}^{T}\bm{u}_{i} + \varepsilon_{i}, \quad i = 1, \dots, n,$$
\end{block}

\begin{itemize}
  \item \small \color{black} $y_{i}$ -- \color{violet} univariate response
  \item \small \color{black} $\bm{x}_{ij} \in \mathbb{R}^{\ell_{j}}$, $j = 1, \dots, p$ -- \color{violet} $p$ subsets of predictors entering indices
  \item \small \color{black} $\bm{\alpha}_{j}$ -- \color{violet} $\ell_{j}$-dimensional vectors of index coefficients
  \item \small \color{black} $g_{j}, f_{k}$ -- \color{violet} smooth nonlinear functions
  \item \small \color{black} Additional predictors :
    \begin{itemize}
      \item \small $w_{ik}$  --   \color{violet} nonlinear
      \item \small \color{black} $\bm{u}_{i}$  --   \color{violet} linear
    \end{itemize}
\end{itemize}


## Sparse multiple index (SMI) model

\begin{block}{}
$$y_{i} = \beta_{0} + \sum_{j = 1}^{p}g_{j}(\bm{\alpha}_{j}^{T}\bm{x}_{ij}) + \sum_{k = 1}^{d}f_{k}(w_{ik}) + \bm{\theta}^{T}\bm{u}_{i} + \varepsilon_{i}, \quad i = 1, \dots, n,$$
\end{block}

\begin{itemize}
  \item \small \color{black} $y_{i}$ -- \color{violet} univariate response
  \item \small \color{black} $\bm{x}_{ij} \in \mathbb{R}^{\ell_{j}}$, $j = 1, \dots, p$ -- \color{violet} $p$ subsets of predictors entering indices
  \item \small \color{black} $\bm{\alpha}_{j}$ -- \color{violet} $\ell_{j}$-dimensional vectors of index coefficients
  \item \small \color{black} $g_{j}, f_{k}$ -- \color{violet} smooth nonlinear functions
  \item \small \color{black} Additional predictors :
    \begin{itemize}
      \item \small $w_{ik}$  --   \color{violet} nonlinear 
      \item \small \color{black} $\bm{u}_{i}$  --   \color{violet} linear
    \end{itemize}
\end{itemize}

\begin{textblock}{5.5}(9, 6.3)
\fontsize{11}{12}\sf
\begin{alertblock}{}
\small Allow elements equal to zero in $\bm{\alpha}_{j}$ -- "Sparse"
\end{alertblock}
\end{textblock}


## Sparse multiple index (SMI) model

\begin{block}{}
$$y_{i} = \beta_{0} + \sum_{j = 1}^{p}g_{j}(\bm{\alpha}_{j}^{T}\bm{x}_{ij}) + \sum_{k = 1}^{d}f_{k}(w_{ik}) + \bm{\theta}^{T}\bm{u}_{i} + \varepsilon_{i}, \quad i = 1, \dots, n,$$
\end{block}

\begin{itemize}
  \item \small \color{black} $y_{i}$ -- \color{violet} univariate response
  \item \small \color{black} $\bm{x}_{ij} \in \mathbb{R}^{\ell_{j}}$, $j = 1, \dots, p$ -- \color{violet} $p$ subsets of predictors entering indices
  \item \small \color{black} $\bm{\alpha}_{j}$ -- \color{violet} $\ell_{j}$-dimensional vectors of index coefficients
  \item \small \color{black} $g_{j}, f_{k}$ -- \color{violet} smooth nonlinear functions
  \item \small \color{black} Additional predictors :
    \begin{itemize}
      \item \small $w_{ik}$  --   \color{violet} nonlinear 
      \item \small \color{black} $\bm{u}_{i}$  --   \color{violet} linear
    \end{itemize}
\end{itemize}

\begin{textblock}{5.5}(9, 6)
\fontsize{11}{12}\sf
\begin{alertblock}{}
\scriptsize Both "p" and the predictor grouping among indices are unknown.
\end{alertblock}
\end{textblock} 

\begin{textblock}{5.5}(9, 7)
\fontsize{11}{12}\sf
\begin{alertblock}{}
\scriptsize Overlapping of predictors among indices is not allowed.
\end{alertblock}
\end{textblock} 


## Benchmarks

## Forecast uncertainty

\begin{itemize}
  \item Uncertainty of a forecast \alert{${\rightarrow}$ Prediction Interval (PI)}
  \pause
  \item Theoretical $100(1 - \alpha)\%$ prediction interval:
$$
  \hat{y}_{t+h|t} \pm z_{\alpha/2} \times \hat{\sigma}_{h},
$$
where
  \begin{itemize}
    \item \small \color{black} $y$ -- \color{violet} time series $y_{1}, \dots, y_{T}$
    \item \small \color{black} $\hat{y}_{t+h|t}$ -- \color{violet} $h$-step-ahead point forecast for $y_{t+h}$ given observations up to $t$
    \item \small \color{black} $z_{\alpha/2}$ -- \color{violet} $\alpha/2$ quantile of standard normal distribution
    \item \small \color{black} $\hat{\sigma}_{h}$ -- \color{violet} estimate of std. deviation of $h$-step forecast distribution
  \end{itemize}
  \pause
  \item Main issue:
  \begin{itemize}
    \item \small \color{blue} Difficult to analytically calculate $h$-step forecast variances for $h > 1$
  \end{itemize}
\end{itemize}


## Block bootstrap

\placefig{1.8}{2.5}{width=12cm}{bbsplit}

## Conformal prediction

\placefig{1.8}{2.5}{width=12cm}{cpsplit}

## Conformal bootstrap


<!-- ## Background -->

<!-- \begin{block}{Block Bootstrap} -->
<!-- \begin{itemize} -->
<!--   \item Resampling from empirical distribution of historical model residuals \color{violet} $\mathbfit{\rightarrow}$ Bootstrapping \newline -->
<!--   \pause -->
<!--   \item \color{black} Randomly resample blocks from the historical model residuals, and join together \alert{$\mathbfit{\rightarrow}$ Block Bootstrapping}  -->
<!--   \item Retains serial correlation in the data  -->
<!--   \pause -->
<!--   \item \color{violet} block length: -->
<!--   \begin{itemize} -->
<!--     \item Long enough to capture autocorrelation patterns -->
<!--     \item Short enough to construct sufficient number of blocks -->
<!--   \end{itemize} -->
<!-- \end{itemize} -->
<!-- \end{block} -->


<!-- ## Background -->

<!-- \begin{block}{\color{yellow} Conformal Prediction (CP) -- Vovk et al. (2005)} -->
<!-- \begin{itemize} -->
<!--   \item A distribution-free approach -->
<!--   \item Relies only on the assumption of \textbf{exchangeability of data} -->
<!--   \item Provides theoretical coverage guarantees -->
<!-- \end{itemize} -->
<!-- \end{block} -->

<!-- \pause -->

<!-- \begin{block}{Split Conformal Prediction (SCP)} -->
<!-- \begin{itemize} -->
<!--   \item A holdout method for generating prediction intervals -->
<!--   \begin{itemize} -->
<!--   \item \textbf{Training set} -- forecasting model is trained -->
<!--   \item \textbf{Calibration set} -- forecasting errors (\textit{nonconformity scores}) are calculated -->
<!--   \item \textbf{Test set} -- prediction intervals are obtained -->
<!--   \end{itemize} -->
<!-- \end{itemize} -->
<!-- \end{block} -->


<!-- ## Background -->

<!-- \begin{itemize} -->
<!--   \item \color{violet}CP methods for \textbf{non-exchangeable data}: -->
<!-- \end{itemize} -->
<!-- \pause -->
<!-- \begin{block}{Weighted Conformal Prediction (WCP) Methods} -->
<!-- \begin{itemize} -->
<!--   \item \color{violet}Tibshirani et al. (2019): -->
<!--   \begin{itemize} -->
<!--     \item Depends on \textbf{"covariate shift"} assumption -->
<!--     \item \textit{Nonconformity scores} are weighted using ratio of likelihoods of training and test covariate distributions -->
<!--     \item Likelihood ratio is assumed to be known or accurately estimated -->
<!--   \end{itemize} -->
<!--   \pause -->
<!--   \item \color{violet}Barber et al. (2023): -->
<!--   \begin{itemize} -->
<!--     \item Weighting \textit{quantiles} to avoid assumption of exchangeability -->
<!--     \item Weights are "fixed" rather than being data dependent -->
<!--   \end{itemize} -->
<!-- \end{itemize} -->
<!-- \end{block} -->


<!-- ## Background -->

<!-- \begin{itemize} -->
<!--   \item \color{violet}CP methods for \textbf{non-exchangeable data}: -->
<!-- \end{itemize} -->
<!-- \begin{block}{Adaptive Conformal Prediction (ACP) -- Gibbs \& Candès (2021)} -->
<!-- \begin{itemize} -->
<!--   \item Update nominal $\mathbfit{\alpha}$ based on achieved coverage -->
<!--   \begin{itemize} -->
<!--     \item If achieved coverage is larger -- increase $\mathbfit{\alpha}$ -->
<!--     \item If achieved coverage is smaller -- decrease $\mathbfit{\alpha}$ -->
<!--   \end{itemize} -->
<!-- \end{itemize} -->
<!-- \end{block} -->


<!-- ## Methodology -->

<!-- \begin{block}{} -->
<!--   \begin{itemize} -->
<!--     \item Prediction interval construction methods -->
<!--     \begin{itemize} -->
<!--       \item Block Bootstrapping (BB) -->
<!--       \item Conformal Prediction (CP) methods: SCP, WSCP, ACP \newline -->
<!--     \end{itemize} -->
<!--     \item Applied using \textit{\color{violet}online learning framework} proposed by \textit{\color{violet}Wang and Hyndman (2024)} \newline -->
<!--   \end{itemize} -->
<!-- \end{block} -->
<!-- \pause -->
<!-- \begin{block}{} -->
<!--   \begin{itemize} -->
<!--     \item We propose a novel methodology \alert{Conformal Block Bootstrap (CBB)} -->
<!--     \begin{itemize} -->
<!--       \item \textbf{\color{blue}A natural integration of BB and SCP} -->
<!--       \item \textbf{\color{blue}Exploits the strengths of both the methods} -->
<!--     \end{itemize} -->
<!--   \end{itemize} -->
<!-- \end{block} -->


<!-- ## Methodology -->

<!-- \fontsize{11}{12}\sf -->
<!-- \begin{itemize}{} -->
<!--   \item \textbf{\color{violet} Block Bootstrap (BB) with Time Series Cross-validation Forecasting} -->
<!-- \end{itemize} -->
<!-- \pause -->
<!-- \begin{block}{} -->
<!-- \fontsize{9}{9}\sf -->
<!-- \color{blue} \textbf{Step 1 :} \color{black} Split data into an initial training window $\{\mathbfit{z}_{1}, \dots, \mathbfit{z}_{w}\}$ and test set $\{\mathbfit{z}_{w + 1}, \dots, \mathbfit{z}_{T}\}$ -->
<!-- \end{block} -->
<!-- \pause -->
<!-- \begin{block}{} -->
<!-- \fontsize{9}{9}\sf -->
<!-- \color{violet} \textbf{Step 2 :} \color{black} Train forecasting model on training set -->
<!-- \end{block} -->
<!-- \pause -->
<!-- \begin{block}{} -->
<!-- \fontsize{9}{9}\sf -->
<!-- \color{blue} \textbf{Step 3 :} \color{black} Obtain series of in-sample residuals $e_{1}, \dots, e_{w}$ -->
<!-- \end{block} -->
<!-- \pause -->
<!-- \begin{block}{} -->
<!-- \fontsize{9}{9}\sf -->
<!-- \color{violet} \textbf{Step 4 :} \color{black} Perform BB by using residuals series; generate several bootstrapped series -->
<!-- \end{block} -->
<!-- \pause -->
<!-- \begin{block}{} -->
<!-- \fontsize{9}{9}\sf -->
<!-- \color{blue} \textbf{Step 5 :} \color{black} Obtain $H$-step-ahead simulated future values, $\hat{y}_{w + 1 \mid w}, \dots, \hat{y}_{w + H \mid w}$, for each bootstrapped series -->
<!-- \end{block} -->
<!-- \pause -->
<!-- \begin{block}{} -->
<!-- \fontsize{9}{9}\sf -->
<!-- \color{violet} \textbf{Step 6 :} \color{black} Calculate quantiles of the sets of simulated future values at each horizon -->
<!-- \end{block} -->
<!-- \pause -->
<!-- \begin{block}{} -->
<!-- \fontsize{9}{9}\sf -->
<!-- \color{blue} \textbf{Step 7 :} \color{black} Iteratively roll training set forward by one observation; repeat steps 2-6 until prediction intervals have been constructed for the entire test set -->
<!-- \end{block} -->






## Forecasting heat exposure-related daily mortality

```{r}
#| label: heat-summer-plot
#| echo: false
#| message: false

dataSummer <- readRDS(here::here("data/Heat_Corrected.rds"))
dataSummer |>
  as_tsibble(index = Date) |>
  mutate(Day = row_number()) |>
  update_tsibble(index = Day, regular = TRUE) |>
  autoplot(Death_lag_000, colour = "#D55E00") +
  scale_x_continuous(
    breaks = seq(1, 2300, by = 92),
    labels = unique(dataSummer$Year)
  ) +
  labs(x = "Date", y = "Number of Deaths", title = "Daily Deaths in Summer - Montreal, Canada") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 12, size = 7))
```


## Forecasting heat exposure-related daily mortality

\begin{textblock}{7}(0.7, 1.5)
\fontsize{11}{12}\sf
\begin{block}{Data}
  \begin{itemize}
    \item \color{violet} \textbf{Response:} \color{black} \textbf{Daily deaths in Summer} -- 1990 to 2014 -- Montreal, Canada
    \item \color{violet} \textbf{Index Variables:} 
      \begin{itemize}
        \item \color{black} Death lags
        \item \color{black} Max temperature lags
        \item \color{black} Min temperature lags
        \item \color{black} Vapor pressure lags
      \end{itemize}
    \item \color{violet}\textbf{Nonlinear:} \color{black} DOS (day of the season), Year \newline
  \end{itemize}
\end{block}
\end{textblock}

## Forecasting heat exposure-related daily mortality

\begin{textblock}{7}(0.7, 1.5)
\fontsize{11}{12}\sf
\begin{block}{Data}
  \begin{itemize}
    \item \color{violet} \textbf{Response:} \color{black} \textbf{Daily deaths in Summer} -- 1990 to 2014 -- Montreal, Canada
    \item \color{violet} \textbf{Index Variables:} 
      \begin{itemize}
        \item \color{black} Death lags
        \item \color{black} Max temperature lags
        \item \color{black} Min temperature lags
        \item \color{black} Vapor pressure lags
      \end{itemize}
    \item \color{violet}\textbf{Nonlinear:} \color{black} DOS (day of the season), Year \newline
  \end{itemize}
\end{block}
\end{textblock}

\begin{textblock}{7}(8.3, 1.5)
\fontsize{11}{12}\sf
\begin{block}{Data split}
  \begin{itemize}
  \item \color{violet}\textbf{Training Set:} \color{black}1990 to 2007 \newline
  \item \color{violet}\textbf{Validation Set:} \color{black}2008 \newline
  \item \color{violet}\textbf{Test Set:} \color{black}2009 to 2014 \newline \newline \newline \newline \newline \newline
\end{itemize}
\end{block}
\end{textblock}


## Conclusion

::: {.callout-note}
## \color{violet} Summary of Results (work-in-progress):

\begin{itemize}
    \item \textbf{Block Bootstrap} -- Under-coverage; too narrow
    \item \textbf{Conformal Prediction} -- Better achieves a target coverage, with acceptable sharpness 
  \end{itemize}
:::

::: {.callout-warning}
## \color{violet} Limitations:

\begin{itemize}
    \item \small Test set is not long enough for larger forecast horizons
    \item \small Hyper-parameter choices
  \end{itemize}
:::



## 

\placefig{1.8}{2.5}{width=12cm}{Findme}



