---
title: Uncertainty Estimation for High-dimensional Nonparametric Forecasting
author: Nuwani Palihawadana
date: 01 July 2025
titlegraphic: bg-13.png
titlecolor: white
toc: false
format:
  presentation-beamer:
    pdf-engine: pdflatex
    keep-tex: true
    fig-width: 8
    fig-height: 4.3
    template-partials:
      - before-title.tex
highlight-style: tango
execute:
  echo: false
  message: false
  warning: false
  cache: false
---

```{r}
#| label: load-packages
#| echo: false
library(knitr)
library(ggplot2)
library(dplyr)
library(tidyr)
library(tsibble)
library(lubridate)
library(cowplot)
library(kableExtra)
library(feasts)
library(patchwork)
```

## Nonparametric forecasting

\placefig{0.5}{1.4}{width=15cm}{nonpara1}


## Nonparametric forecasting

\placefig{0.5}{1.3}{width=15cm}{nonpara2}


## Sparse Multiple Index (SMI) model

\placefig{0.5}{1.5}{width=15cm}{smi1}

## Sparse Multiple Index (SMI) model

\placefig{0.5}{1.5}{width=15cm}{smi2}

## Sparse Multiple Index (SMI) model

\placefig{0.5}{1.5}{width=15cm}{smi3}


## Other models

\begin{itemize}
  \item \color{violet} Nonparametric additive model with backward elimination (Backward):
  \begin{itemize}
    \item No linear combinations (indices)
    \item Fully additive \newline
  \end{itemize}
  \pause
  \item \color{violet} Groupwise Additive Index Model (GAIM):
  \begin{itemize}
    \item Predefined predictor groups
    \item No overlapping predictors among groups \newline
  \end{itemize}
  \pause
  \item \color{violet} Projection Pursuit Regression model (PPR):
  \begin{itemize}
    \item All predictors enter all indices
  \end{itemize}
\end{itemize}


## Forecast uncertainty

\begin{itemize}
  \item Uncertainty of a forecast \alert{${\rightarrow}$ Prediction Interval (PI)}
  \pause
  \item Theoretical $100(1 - \alpha)\%$ prediction interval:
$$
  \hat{y}_{t+h|t} \pm z_{\alpha/2} \times \hat{\sigma}_{h},
$$
where
  \begin{itemize}
    \item \small \color{black} $y$ -- \color{violet} time series $y_{1}, \dots, y_{T}$
    \item \small \color{black} $\hat{y}_{t+h|t}$ -- \color{violet} $h$-step-ahead point forecast for $y_{t+h}$ given observations up to $t$
    \item \small \color{black} $z_{\alpha/2}$ -- \color{violet} $\alpha/2$ quantile of standard normal distribution
    \item \small \color{black} $\hat{\sigma}_{h}$ -- \color{violet} estimate of std. deviation of $h$-step forecast distribution
  \end{itemize}
  \pause
  \item Main issue:
  \begin{itemize}
    \item \small \color{blue} Difficult to analytically calculate $h$-step forecast variances for $h > 1$
  \end{itemize}
\end{itemize}


## Block bootstrap (BB)

\alert{Block bootstrapping} \newline

\begin{itemize}
   \item Randomly resample blocks from the historical model residuals, and join together \newline
  \item Retains serial correlation in the data
\end{itemize}


## BB with time series cross-validation Forecasting

<!-- \fontsize{11}{12}\sf -->
<!-- \begin{itemize}{} -->
<!--   \item \textbf{\color{violet} Block bootstrap with time series cross-validation Forecasting} -->
<!-- \end{itemize} -->

\begin{block}{}
\fontsize{9}{9}\sf
\color{blue} \textbf{Step 1 :} \color{black} Split the data into an initial training window, $\{\bm{z}_{1}, \dots, \bm{z}_{tr}\}$, and an initial test window, $\{\bm{z}_{tr + 1}, \dots, \bm{z}_{tr + H}\}$.
\end{block}

\begin{block}{}
\fontsize{9}{9}\sf
\color{violet} \textbf{Step 2 :} \color{black} Train forecasting model on training window.
\end{block}

\begin{block}{}
\fontsize{9}{9}\sf
\color{blue} \textbf{Step 3 :} \color{black} Obtain the series of in-sample residuals, $e_{1}, \dots, e_{tr}$.
\end{block}

\begin{block}{}
\fontsize{9}{9}\sf
\color{violet} \textbf{Step 4 :} \color{black} Perform BB by using residuals series; generate several bootstrapped series.
\end{block}

\begin{block}{}
\fontsize{9}{9}\sf
\color{blue} \textbf{Step 5 :} \color{black} Obtain $H$-step-ahead simulated future values, $\hat{y}_{tr + 1 \mid tr}, \dots, \hat{y}_{tr + H \mid tr}$, for each bootstrapped series.
\end{block}

\begin{block}{}
\fontsize{9}{9}\sf
\color{violet} \textbf{Step 6 :} \color{black} Calculate quantiles of the sets of simulated future values at each horizon.
\end{block}

\begin{block}{}
\fontsize{9}{9}\sf
\color{blue} \textbf{Step 7 :} \color{black} Iteratively roll training window forward by one observation; repeat steps 2--6 until prediction intervals have been constructed for the entire test set.
\end{block}


## BB with time series cross-validation Forecasting

\placefig{1}{2.5}{width=10cm}{bbsplit}

\begin{textblock}{4}(11.5, 3)
\fontsize{11}{12}\sf
\begin{block}{}
  \begin{itemize}
  \item \color{violet} {$tr$} : \color{black} length of training window \newline
  \item \color{violet} {$H$} : \color{black} forecast horizon \newline
\end{itemize}
\end{block}
\end{textblock}


## Conformal prediction (CP)


\alert{Conformal prediction} \newline
\begin{itemize}
  \item A distribution-free approach (Vovk et al. 2005) 
  \item Relies only on the assumption of \textbf{exchangeability of data} 
  \item Provides theoretical coverage guarantees \newline
\end{itemize}
\pause
 \color{violet}\textbf{Split Conformal Prediction (SCP) :}
\begin{itemize}
  \item A holdout method for generating prediction intervals
  \begin{itemize}
  \item \textbf{Training set} -- train forecasting model
  \item \textbf{Calibration set} -- calculate forecast errors (\textit{nonconformity scores}) 
  \item \textbf{Test set} -- obtain prediction intervals
  \end{itemize}
\end{itemize}


## CP methods for non-exchangeable data

\color{violet}\textbf{Weighted Split Conformal Prediction (WSCP) (Barber et al. 2023) :}
\begin{itemize}
  \item Weighting quantiles using fixed (data-independent) weights \newline
\end{itemize}
\pause
\color{violet}\textbf{Adaptive Conformal Prediction (ACP) (Gibbs \& Candès 2021) :}
\begin{itemize}
  \item Update nominal $\bm{\alpha}$ based on achieved coverage
\end{itemize}


## CP with time series cross-validation Forecasting

\begin{block}{}
\fontsize{9}{9}\sf
\color{blue} \textbf{Step 1 :} \color{black} Split data into an initial training window, $\{\bm{z}_{1}, \dots, \bm{z}_{tr}\}$, an initial calibration window, $\{\bm{z}_{tr + 1}, \dots, \bm{z}_{tr + tc}\}$, and an initial test window, $\{\bm{z}_{tr + tc + 1}, \dots, \bm{z}_{tr + tc + H}\}$.
\end{block}

\begin{block}{}
\fontsize{9}{9}\sf
\color{violet} \textbf{Step 2 :} \color{black} Estimate forecasting model on initial training window. Obtain $H$-step-ahead forecasts, $\hat{y}_{tr + h \mid tr}$, and forecast errors $s_{tr + h \mid tr} = y_{tr + h} - \hat{y}_{tr + h \mid tr}$, for $h = 1, \dots, H$.
\end{block}

\begin{block}{}
\fontsize{9}{9}\sf
\color{blue} \textbf{Step 3 :} \color{black} Perform cross-validation forecasting, while repeating step 2, until forecast errors are obtained for entire initial calibration window.
\end{block}

\begin{block}{}
\fontsize{9}{9}\sf
\color{violet} \textbf{Step 4 :} \color{black} Apply the conformal prediction method of interest: SCP, WSCP, or ACP, to compute $H$-step prediction intervals.
\end{block}

\begin{block}{}
\fontsize{9}{9}\sf
\color{blue} \textbf{Step 5 :} \color{black} Iteratively roll training window and calibration window forward by one observation; repeat steps 2--4 until prediction intervals are obtained for the entire test set. 
\end{block}


## CP with time series cross-validation Forecasting

\placefig{1}{2.5}{width=10cm}{cpsplit}

\begin{textblock}{4}(11.5, 2.2)
\fontsize{11}{12}\sf
\begin{block}{}
  \begin{itemize}
  \item \color{violet} {$tr$} : \color{black} length of training window \newline
  \item \color{violet} {$tc$} : \color{black} length of calibration window \newline
  \item \color{violet} {$H$} : \color{black} forecast horizon \newline
\end{itemize}
\end{block}
\end{textblock}


## Conformal bootstrap (CB)


\alert{Conformal bootstrap}
\begin{itemize}
  \item \textbf{A natural integration of BB and SCP} \newline
  \pause
  \item \textbf{Exploits the strengths of both the methods} \newline
  \begin{itemize}
  \item \color{violet} Preserves temporal dependencies inherent in multi-step forecasts \newline
  \item \color{violet} Accounts for additional uncertainty brought into the process by ex-ante forecasting
\end{itemize}
\end{itemize}


## CB with time series cross-validation Forecasting

\placefig{1}{2.5}{width=10cm}{cpsplit}

\begin{textblock}{4}(11.5, 2.2)
\fontsize{11}{12}\sf
\begin{block}{}
  \begin{itemize}
  \item \color{violet} {$tr$} : \color{black} length of training window \newline
  \item \color{violet} {$tc$} : \color{black} length of calibration window \newline
  \item \color{violet} {$H$} : \color{black} forecast horizon \newline
\end{itemize}
\end{block}
\end{textblock}


## CB with time series cross-validation Forecasting

\begin{block}{}
\fontsize{9}{9}\sf
\color{blue} \textbf{Step 1 :} \color{black} Split data into an initial training window, $\{\bm{z}_{1}, \dots, \bm{z}_{tr}\}$, an initial calibration window, $\{\bm{z}_{tr + 1}, \dots, \bm{z}_{tr + tc}\}$, and an initial test window, $\{\bm{z}_{tr + tc + 1}, \dots, \bm{z}_{tr + tc + H}\}$.
\end{block}

\pause

\begin{block}{}
\fontsize{9}{9}\sf
\color{violet} \textbf{Step 2 :} \color{black} Estimate forecasting model on initial training window. Obtain $H$-step-ahead forecasts, $\hat{y}_{tr + h \mid tr}$, and forecast errors $s_{tr + h \mid tr} = y_{tr + h} - \hat{y}_{tr + h \mid tr}$, for $h = 1, \dots, H$.
\end{block}

\pause

\begin{block}{}
\fontsize{9}{9}\sf
\color{blue} \textbf{Step 3 :} \color{black} Perform cross-validation forecasting, while repeating step 2, until forecast errors are obtained for entire initial calibration window.
\end{block}

\pause

\begin{block}{}
\fontsize{9}{9}\sf
\color{violet} \textbf{Step 4 :} \color{black} Use the model estimated on the most recent training window to generate $H$-step-ahead simulated future values.
\begin{itemize}
  \item Autoregressive: block bootstrap 1-step forecast errors; block size $ = H$.
  \item Non-autoregressive: bootstrap multi-step forecast errors at each time point as a whole.
\end{itemize}
\end{block}

\pause

\begin{block}{}
\fontsize{9}{9}\sf
\color{blue} \textbf{Step 5 :} \color{black} Calculate quantiles of the sets of simulated future values at each horizon.
\end{block}

\pause

\begin{block}{}
\fontsize{9}{9}\sf
\color{violet} \textbf{Step 6 :} \color{black} Iteratively roll training window forward by one observation; repeat steps 2--5 until prediction intervals have been constructed for the entire test set. 
\end{block}


## Forecasting heat exposure-related daily mortality

```{r}
#| label: heat-summer-plot
#| echo: false
#| message: false

dataSummer <- readRDS(here::here("data/Heat_Corrected.rds"))
dataSummer |>
  as_tsibble(index = Date) |>
  mutate(Day = row_number()) |>
  update_tsibble(index = Day, regular = TRUE) |>
  autoplot(Death_lag_000, colour = "#D55E00") +
  scale_x_continuous(
    breaks = seq(1, 2300, by = 92),
    labels = unique(dataSummer$Year)
  ) +
  labs(x = "Date", y = "Number of Deaths", title = "Daily Deaths in Summer - Montreal, Canada") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 12, size = 7))
```


## Forecasting heat exposure-related daily mortality

\begin{textblock}{7}(0.7, 1.5)
\fontsize{11}{12}\sf
\begin{block}{Data}
  \begin{itemize}
    \item \color{violet} \textbf{Response:} \color{black} \textbf{Daily deaths in Summer} -- 1990 to 2014 -- Montreal, Canada
    \item \color{violet} \textbf{Index Variables:} 
      \begin{itemize}
        \item \color{black} Death lags
        \item \color{black} Max temperature lags
        \item \color{black} Min temperature lags
        \item \color{black} Vapor pressure lags
      \end{itemize}
    \item \color{violet}\textbf{Nonlinear:} \color{black} DOS (day of the season), Year \newline
  \end{itemize}
\end{block}
\end{textblock}

## Forecasting heat exposure-related daily mortality

\begin{textblock}{7}(0.7, 1.5)
\fontsize{11}{12}\sf
\begin{block}{Data}
  \begin{itemize}
    \item \color{violet} \textbf{Response:} \color{black} \textbf{Daily deaths in Summer} -- 1990 to 2014 -- Montreal, Canada
    \item \color{violet} \textbf{Index Variables:} 
      \begin{itemize}
        \item \color{black} Death lags
        \item \color{black} Max temperature lags
        \item \color{black} Min temperature lags
        \item \color{black} Vapor pressure lags
      \end{itemize}
    \item \color{violet}\textbf{Nonlinear:} \color{black} DOS (day of the season), Year \newline
  \end{itemize}
\end{block}
\end{textblock}

\begin{textblock}{7}(8.3, 1.5)
\fontsize{11}{12}\sf
\begin{block}{Data split}
  \begin{itemize}
  \item \color{violet} {$\bm{tr}$ - training window} : \color{black} 1748 \newline
  \item \color{violet} {$\bm{tc}$ - calibration window} : \color{black} 300 \newline
  \item \color{violet} {$\bm{H}$ - forecast horizon} : \color{black} 92 \newline \newline \newline \newline \newline
\end{itemize}
\end{block}
\end{textblock}


## Mean coverage of $95\%$ prediction intervals

```{r}
#| echo: false
#| message: false

heat_cov_data_smimodel_ppr <- readRDS(here::here("results/heat_smimodel_ppr_cov_plotdata_combined_chp3_smi.rds"))
heat_cov_data_backward <- readRDS(here::here("results/heat_backward_cov_plotdata_combined_chp3_final_corrected.rds"))
heat_cov_data_gaim <- readRDS(here::here("results/heat_gaim_cov_plotdata_combined_chp3_final_corrected.rds"))
heat_cov_data_ppr <- readRDS(here::here("results/heat_ppr_cov_plotdata_combined_chp3_final_corrected.rds"))
methods <- c("BB", "SCP", "WSCP", "ACP", "CB")
cols <- c(
  "BB" = "#CC79A7",
  "SCP" = "#0072B2",
  "WSCP" = "#009E73",
  "ACP" = "#D55E00",
  "CB" = "black"
)

p1 <- heat_cov_data_smimodel_ppr |>
  mutate(method = factor(method, levels = methods)) |>
  ggplot(aes(x = horizon, y = coverage, group = method, colour = method)) +
  geom_line(linewidth = 0.4, alpha = 0.8) +
  scale_colour_manual(values = cols) +
  geom_hline(yintercept = 0.95, linetype = "dashed", colour = "gray") +
  xlab("Forecast horizon") +
  ylab("Mean coverage") +
  ylim(c(0.75, 1.00)) +
  ggtitle("SMI") +
  theme_bw() +
  theme(axis.title.x = element_blank()) +
  scale_x_discrete(breaks = seq(1, 92, by = 5))

p2 <- heat_cov_data_backward |>
  mutate(method = factor(method, levels = methods)) |>
  ggplot(aes(x = horizon, y = coverage, group = method, colour = method)) +
  geom_line(linewidth = 0.4, alpha = 0.8) +
  scale_colour_manual(values = cols) +
  geom_hline(yintercept = 0.95, linetype = "dashed", colour = "gray") +
  xlab("Forecast horizon") +
  ylab("Mean coverage") +
  ylim(c(0.75, 1.00)) +
  ggtitle("Backward") +
  theme_bw() +
  theme(axis.title.x = element_blank(),
        axis.title.y = element_blank()) +
  scale_x_discrete(breaks = seq(1, 92, by = 5))

p3 <- heat_cov_data_gaim |>
  mutate(method = factor(method, levels = methods)) |>
  ggplot(aes(x = horizon, y = coverage, group = method, colour = method)) +
  geom_line(linewidth = 0.4, alpha = 0.8) +
  scale_colour_manual(values = cols) +
  geom_hline(yintercept = 0.95, linetype = "dashed", colour = "gray") +
  xlab("Forecast horizon") +
  ylab("Mean coverage") +
  ylim(c(0.75, 1.00)) +
  ggtitle("GAIM") +
  theme_bw() +
  scale_x_discrete(breaks = seq(1, 92, by = 5))

p4 <- heat_cov_data_ppr |>
  mutate(method = factor(method, levels = methods)) |>
  ggplot(aes(x = horizon, y = coverage, group = method, colour = method)) +
  geom_line(linewidth = 0.4, alpha = 0.8) +
  scale_colour_manual(values = cols) +
  geom_hline(yintercept = 0.95, linetype = "dashed", colour = "gray") +
  xlab("Forecast horizon") +
  ylab("Mean coverage") +
  ylim(c(0.75, 1.00)) +
  ggtitle("PPR") +
  theme_bw() +
  theme(axis.title.y = element_blank()) +
  scale_x_discrete(breaks = seq(1, 92, by = 5))

patchwork <- (p1 + p2) / (p3 + p4) & theme(legend.position = "bottom", 
                                           #legend.box = "vertical",
                                           legend.title = element_blank(),
                                           axis.text.x = element_text(size = 7),
                                           axis.title = element_text(size = 8),
                                           title = element_text(size = 8))
patchwork +
  plot_layout(guides = "collect")

```


## Mean width of $95\%$ prediction intervals

```{r}
#| echo: false
#| message: false

heat_width_data_smimodel_ppr <- readRDS(here::here("results/heat_smimodel_ppr_width_plotdata_combined_chp3_smi.rds"))
heat_width_data_backward <- readRDS(here::here("results/heat_backward_width_plotdata_combined_chp3_final_corrected.rds"))
heat_width_data_gaim <- readRDS(here::here("results/heat_gaim_width_plotdata_combined_chp3_final_corrected.rds"))
heat_width_data_ppr <- readRDS(here::here("results/heat_ppr_width_plotdata_combined_chp3_final_corrected.rds"))

p5 <- heat_width_data_smimodel_ppr |>
  mutate(method = factor(method, levels = methods)) |>
  ggplot(aes(x = horizon, y = width, group = method, colour = method)) +
  geom_line(linewidth = 0.4, alpha = 0.8) +
  scale_colour_manual(values = cols) +
  xlab("Forecast horizon") +
  ylab("Mean width") +
  ylim(c(30, 80)) +
  ggtitle("SMI-P") +
  theme_bw() +
  theme(axis.title.x = element_blank()) +
  scale_x_discrete(breaks = seq(1, 92, by = 5))

p6 <- heat_width_data_backward |>
  mutate(method = factor(method, levels = methods)) |>
  ggplot(aes(x = horizon, y = width, group = method, colour = method)) +
  geom_line(linewidth = 0.4, alpha = 0.8) +
  scale_colour_manual(values = cols) +
  xlab("Forecast horizon") +
  ylab("Mean width") +
  ylim(c(30, 80)) +
  ggtitle("Backward") +
  theme_bw() +
  theme(axis.title.x = element_blank(),
        axis.title.y = element_blank()) +
  scale_x_discrete(breaks = seq(1, 92, by = 5))

p7 <- heat_width_data_gaim |>
  mutate(method = factor(method, levels = methods)) |>
  ggplot(aes(x = horizon, y = width, group = method, colour = method)) +
  geom_line(linewidth = 0.4, alpha = 0.8) +
  scale_colour_manual(values = cols) +
  xlab("Forecast horizon") +
  ylab("Mean width") +
  ylim(c(30, 80)) +
  ggtitle("GAIM") +
  theme_bw() +
  scale_x_discrete(breaks = seq(1, 92, by = 5))

p8 <- heat_width_data_ppr |>
  mutate(method = factor(method, levels = methods)) |>
  ggplot(aes(x = horizon, y = width, group = method, colour = method)) +
  geom_line(linewidth = 0.4, alpha = 0.8) +
  scale_colour_manual(values = cols) +
  xlab("Forecast horizon") +
  ylab("Mean width") +
  ylim(c(30, 80)) +
  ggtitle("PPR") +
  theme_bw() +
  theme(axis.title.y = element_blank()) +
  scale_x_discrete(breaks = seq(1, 92, by = 5))

patchwork2 <- (p5 + p6) / (p7 + p8) & theme(legend.position = "bottom", 
                                           #legend.box = "vertical",
                                           legend.title = element_blank(),
                                           axis.text.x = element_text(size = 7),
                                           axis.title = element_text(size = 8),
                                           title = element_text(size = 8))
patchwork2 +
  plot_layout(guides = "collect")
```


## Conclusion

::: {.callout-note}
## \color{blue} Summary of results:

\begin{itemize}
    \item \textbf{\color{violet} Block bootstrap} prediction intervals were \textbf{narrow} and resulted in \textbf{under-coverage}
    \pause
    \item \textbf{\color{violet} Conformal prediction} methods \textbf{generally performed well} across models and forecast horizons
    \pause
    \item \textbf{\color{violet} Conformal bootstrap} method \textbf{performed well in shorter horizons}; performance decreased in longer horizons
    \begin{itemize}
      \pause
      \item \color{brown} Using only 1-step errors
      \pause
      \item \color{brown} Predictor truncation when simulating sample paths
    \end{itemize}
    \pause
    \item \textbf{\color{violet}Coverage performance} is influenced by: \textbf{data characteristics, forecasting model}, and \textbf{forecast horizon}
\end{itemize}
:::


## R Package - smimodel

\placefig{11.7}{2.2}{width=4cm}{hex-large}

\begin{textblock}{10.2}(1, 2.2)
\fontsize{12}{14}\sf
\begin{block}{}
  \begin{itemize}
    \item Block bootstrap
    \begin{itemize}
      \item \color{violet} \textbf{bb\_cvforecast()} \newline
    \end{itemize}
    \item Conformal bootstrap
    \begin{itemize}
      \item \color{violet} \textbf{cb\_cvforecast()} \newline
    \end{itemize}
  \end{itemize}
\end{block}
\end{textblock}

\begin{textblock}{10.2}(1, 6)
\fontsize{11}{12}\sf
\begin{alertblock}{}
 \color{blue} \hspace*{0.8cm} \href{https://github.com/nuwani-palihawadana/smimodel}{github.com/nuwani-palihawadana/smimodel}
\end{alertblock}
\end{textblock}



## 

\placefig{1.8}{2.5}{width=12cm}{Findme}



